---
title: "VAR models"
author: "Ryan Peterson"
date: "2024-05-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library}
#for data/date manipulating
suppressMessages(library(lubridate))
suppressMessages(library(tidyverse))
suppressMessages(library(fields)) #test.for.zero()

#modeling checks
suppressMessages(library(MASS))

#ts data types
suppressMessages(library(xts))
suppressMessages(library(zoo))

#time series 
suppressMessages(library(vars)) #vector autoregression
suppressMessages(library(forecast)) #spectral methods
suppressMessages(library(marima)) #VARIMA
#suppressMessages(library(MTS)) #try mts if needed
```


```{r imported_functions}
#lag_function (requires tidyverse, lubridate)
setwd("~/CO_AUS/Aus_CO-main")
source("lag_function.R")
```


```{r data}
setwd("~/CO_AUS/Aus_CO-main")

#anomaly only response with week data
response_anoms <- read.csv("Data/response_anoms.csv", header = TRUE, stringsAsFactors = FALSE)

#predictors with week data
predictor_anoms <- read.csv("Data/predictor_anoms.csv", header = TRUE, stringsAsFactors = FALSE)

```

Temporary removal of 2019/2020 season in below code.

```{r data_cleaning}
#TODO: potentially create a function for date bounding
#cleaning data for time series work
week <- 1

response <- response_anoms
predictor <- predictor_anoms

resp_dates <- as_date(response[response$week == week, ]$time)
pred_dates <- as_date(predictor[predictor$week == week, ]$time)

#get minimum prediction date from response data
low_bound <- pred_dates[which(pred_dates == min(resp_dates)) - 1]
#get maximum response date from predictor data
upper_bound <- max(as_date(predictor$time)) + days(7)

#this requires week being set correctly above, that req should be fixed
bounded_pred_df <- predictor[predictor$time >= low_bound & predictor$time <= upper_bound, ]

bounded_resp_df <- response[response$time <= upper_bound, ]#temporarily changed so that each week has the same number of years

#2019 boundary (week 14 2019: end of 2018/2019 season)
max_date <- as_date(response_anoms[response_anoms$week == 14 & 
                                     response_anoms$year == 2019, ]$time)

bounded_pred_df <- bounded_pred_df[bounded_pred_df$time <= max_date, ]
bounded_resp_df <- bounded_resp_df[bounded_resp_df$time <= max_date, ]

rm(response, predictor, low_bound, upper_bound, 
   week, resp_dates, pred_dates, max_date)

#outputs to check the boundary dates are correct
min(bounded_pred_df$time)
min(bounded_resp_df$time)

max(bounded_pred_df$time)
max(bounded_resp_df$time)
```

Make more bounding adjusmtents, that we perfomr later on.
That is:
-combine week 53 to week 52
-

```{r further_df_adj}
#var check
temp_week_52 <- bounded_resp_df[bounded_resp_df$week == 52, ]

#correct for week 53
week_53 <- which(bounded_resp_df$week == 53, )
for (k in week_53) {
  bounded_resp_df[k-1, 3] <- mean(c(bounded_resp_df[k-1, 3], 
                                    bounded_resp_df[k, 3]))
  bounded_resp_df[k-1, 4] <- mean(c(bounded_resp_df[k-1, 4], 
                                    bounded_resp_df[k, 4]))
}
rm(k)
bounded_resp_df <- bounded_resp_df[-week_53, ]

#fix for the correct columns
for (j in week_53) {
  bounded_pred_df[j-1, 3] <- mean(c(bounded_pred_df[j-1, 3], 
                                    bounded_pred_df[j, 3]))
  bounded_pred_df[j-1, 4] <- mean(c(bounded_pred_df[j-1, 4],
                                    bounded_pred_df[j, 4]))
  bounded_pred_df[j-1, 5] <- mean(c(bounded_pred_df[j-1, 5],
                                    bounded_pred_df[j, 5]))
  bounded_pred_df[j-1, 6] <- mean(c(bounded_pred_df[j-1, 6], 
                                    bounded_pred_df[j, 6]))
}
rm(j)
bounded_pred_df <- bounded_pred_df[-week_53, ]

#alter the data to align better, then we can make adjustments to this later
min_pred <- min(bounded_resp_df$time)
max_resp <- max(bounded_pred_df$time)

bounded_resp_df <- bounded_resp_df[bounded_resp_df$time <= max_resp, ]
bounded_pred_df <- bounded_pred_df[bounded_pred_df$time >= min_pred, ]

rm(min_pred, max_resp)

```


```{r season_setup}
season_weeks <- c(35:52, 1:14)

#early season weeks 35-42
early <- c(35:42)
#early-mid 43-50
early_mid <- c(43:50)
#early-late 52-6
late_mid <- c(51:52, 1:6)
#late 6-14
late <- c(7:14)
```


Outline/Goals:

1. Build simple var model with a response region (NE/SE AUS) and a single climate mode series, with shorter range dependencies (e.g p <= 20).

2. Find optimal  (via AIC/BIC) VAR model for our response time series of NE Aus and SE Aus.

3. Visualizations:
- CCF plot for each response and each predictor (total 8 figures)
- residual plots
- fitted curve over true data
- maybe ACF/PACF residuals (already done, maybe include in paper/presentation.)


Notes/Comments
- For the var() function use the season argument to identify the wildfire season along with partitions of the wildfire season.
- Add in seasonality by creating another variable that has the data for that season, basically in season and out of season data.
- Use method from MTS to filter out non-significant parameters in a model.

Questions:


# Time Series Checks 


CCF models

```{r}
NEaus_ts <- ts(bounded_resp_df$NE_Aus_anomaly_co, start = as.Date(min(bounded_resp_df$time)), frequency = 52)
SEaus_ts <- ts(bounded_resp_df$SE_Aus_anomaly_co, start = as.Date(min(bounded_resp_df$time)), frequency = 52)

nino_ts <- ts(bounded_pred_df$nino.anomaly, start = as.Date(min(bounded_pred_df$time)), frequency = 52)
dmi_ts <- ts(bounded_pred_df$dmi.anomaly, start = as.Date(min(bounded_pred_df$time)), frequency = 52)
tsa_ts <- ts(bounded_pred_df$tsa.anomaly, start = as.Date(min(bounded_pred_df$time)), frequency = 52)
aao_ts <- ts(bounded_pred_df$aao.anomaly, start = as.Date(min(bounded_pred_df$time)), frequency = 52)

ccf(as.numeric(NEaus_ts), as.numeric(nino_ts), type = "correlation", lag = 52)
ccf(as.numeric(SEaus_ts), as.numeric(nino_ts), type = "correlation", lag = 52)
ccf(as.numeric(NEaus_ts), as.numeric(dmi_ts), type = "correlation", lag = 52)
ccf(as.numeric(SEaus_ts), as.numeric(dmi_ts), type = "correlation", lag = 52)
ccf(as.numeric(NEaus_ts), as.numeric(tsa_ts), type = "correlation", lag = 52)
ccf(as.numeric(SEaus_ts), as.numeric(tsa_ts), type = "correlation", lag = 52)
ccf(as.numeric(NEaus_ts), as.numeric(aao_ts), type = "correlation", lag = 52)
ccf(as.numeric(SEaus_ts), as.numeric(aao_ts), type = "correlation", lag = 52)
```

(variance)

```{r var_adj}
#negative adjustments
y <- bounded_resp_df$NE_Aus_anomaly_co
c <- if(min(y) <= 0) abs(min(y)) + 1.01 else 0
y_pos_NEaus <- log(y + c)

y <- bounded_resp_df$SE_Aus_anomaly_co
c <- if(min(y) <= 0) abs(min(y)) + 1.01 else 0
y_pos_SEaus <- log(y + c)

#new predictor df w/ only pred values
new_pred_df <- bounded_pred_df[ ,3:6]

#fit for box cox
fit_test  <- lm(y_pos_NEaus ~ 1, data = new_pred_df)

boxcox_result <- boxcox(fit_test, lambda = seq(0, 5, by = 0.1))
lambda_optimal <- boxcox_result$x[which.max(boxcox_result$y)]

trans_yposNEaus <- (y_pos_NEaus^lambda_optimal - 1) / lambda_optimal

fit_test  <- lm(y_pos_SEaus ~ 1, data = new_pred_df)

boxcox_result <- boxcox(fit_test, lambda = seq(0, 5, by = 0.1))
lambda_optimal <- boxcox_result$x[which.max(boxcox_result$y)]

trans_yposSEaus <- (y_pos_NEaus^lambda_optimal - 1) / lambda_optimal

```

```{r data_ts}

```


```{r DONT RUN}
#normalizing for both values
#for (i in 1:52) {
#  bounded_resp_df[,i] <- (bounded_resp_df[,i] - mean(bounded_resp_df[,i]))/ sd(bounded_resp_df[,i])
#}


```


```{r variance_plot}
#plot variance wrt time
temp_bounded_resp_df <- bounded_resp_df
temp_bounded_resp_df$NE_Aus_anomaly_co <- trans_yposNEaus
temp_bounded_resp_df$SE_Aus_anomaly_co <- trans_yposSEaus

tempNEaus_ts <- ts(temp_bounded_resp_df$NE_Aus_anomaly_co, 
               start = as.Date(min(temp_bounded_resp_df$time)), frequency = 52)
tempSEaus_ts <- ts(temp_bounded_resp_df$SE_Aus_anomaly_co, 
               start = as.Date(min(temp_bounded_resp_df$time)), frequency = 52)

plot.ts(tempNEaus_ts)
plot.ts(tempSEaus_ts)

#variance by week
aus_var <- as.data.frame(matrix(ncol = 3) )
colnames(aus_var) <- c("NEaus_var", "SEaus_var", "week")
for (i in 1:52) {
  temp_week <- temp_bounded_resp_df[temp_bounded_resp_df$week == i, ]
  temp_row <- c(var(temp_week[,3]), var(temp_week[,4]), i)
  aus_var <- rbind(aus_var, temp_row)
}
aus_var <- aus_var[-1, ]

plot(aus_var$week, aus_var$NEaus_var, type = "l")
plot(aus_var$week, aus_var$SEaus_var, type = "l")
```


```{r hist}
hist(bounded_resp_df$NE_Aus_anomaly_co, breaks = 12, col = 'steelblue', 
     main = "NE AUS, no adj",  xlab = "NE Aus CO Anomalies") 
hist(temp_bounded_resp_df$NE_Aus_anomaly_co, breaks = 12, col = 'steelblue',
     main = "NE AUS, box-cox",  xlab = "NE Aus CO Anomalies")


hist(bounded_resp_df$SE_Aus_anomaly_co, breaks = 12, col = 'steelblue',
     main = "SE AUS, no adj.", xlab = "SE Aus CO Anomalies")
hist(temp_bounded_resp_df$SE_Aus_anomaly_co, breaks = 12, col = 'steelblue',
     main = "SE AUS, box-cox",  xlab = "SE Aus CO Anomalies")
```


# 1. Var models

Considering NE AUS and Nino

```{r var_data_selection}
#TODO: add in se_aus

#clean data a little bit more
ne_aus_df <- bounded_resp_df[ ,c(2,3,5,6)]
nino_df <- bounded_pred_df[ ,c(2,3,7,8)]
dmi_df <- bounded_pred_df[ ,c(2,4,7,8)]
tsa_df <- bounded_pred_df[ ,c(2,5,7,8)]
aao_df <- bounded_pred_df[ ,c(2,6,7,8)]

#alter the data to align better, then we can make adjustments to this later
min_pred <- min(ne_aus_df$time)
max_resp <- max(nino_df$time)

ne_aus_df <- ne_aus_df[ne_aus_df$time <= max_resp, ]
nino_df <- nino_df[nino_df$time >= min_pred, ]
dmi_df <- dmi_df[dmi_df$time >= min_pred, ]
tsa_df <- tsa_df[tsa_df$time >= min_pred, ]
aao_df <- aao_df[aao_df$time >= min_pred, ]

min_date <- as.Date(min(nino_df$time))

#create a df and ts-object
dates <- ne_aus_df$time
NE_aus_tsdf <- data.frame(Date = dates, NE_AUS = ne_aus_df$NE_Aus_anomaly_co,
                          Nino = nino_df$nino.anomaly,
                          dmi = dmi_df$dmi.anomaly,
                          tsa = tsa_df$tsa.anomaly,
                          aao = aao_df$aao.anomaly)

NE_aus_zoo <- zoo(NE_aus_tsdf[, -1], order.by = NE_aus_tsdf$Date)

rm(min_pred, max_resp, min_date)
```


```{r just_test_it_out}
#no seasonality
var_select <- VARselect(NE_aus_zoo, lag.max = 52, type = "both")
aic <- var_select$criteria[1,]
bic <- var_select$criteria[3,]

aic_test <- rep(NA, 10)
for (i in 1:10) {
  NE_temp <- vars::VAR(NE_aus_zoo, p = i, type = "both")
  NE_aus_temp <- NE_temp$varresult$NE_AUS
  aic_test[i] <- AIC(NE_aus_temp, k = i + 2)
}

plot(1:10, aic_test, type = "l", xlab = "p", ylab = "AIC", main = "NE Aus VAR(p)")

NE_model <- vars::VAR(NE_aus_zoo, p = 4, type = "both")

NE_aus_model <- NE_model$varresult$NE_AUS

AIC(NE_aus_model, k =2)

summary(NE_model$varresult$NE_AUS)

#change var names and titles below
residuals_var <- residuals(NE_model)

# Plot ACF and PACF for the first variable's residualsindices 
acf(residuals_var[,1], main="ACF of Residuals for NE Aus")
Pacf(residuals_var[,1], main="PACF of Residuals for NE Aus")

ccf(residuals_var[, 1], residuals_var[, 2], lag.max = 15, main = "Partial CCF Between Residuals of Series 1 and 2")
ccf(residuals_var[, 1], residuals_var[, 3], lag.max = 15, main = "Partial CCF Between Residuals of Series 1 and 3")
ccf(residuals_var[, 1], residuals_var[, 4], lag.max = 15, main = "Partial CCF Between Residuals of Series 1 and 4")
ccf(residuals_var[, 1], residuals_var[, 5], lag.max = 15, main = "Partial CCF Between Residuals of Series 1 and 5")

plot(1:52, aic, type = "l", xlab = "lag")
plot(1:52, bic, type = "l", xlab = "lag")
```

```{r quick_varma}
model <- define.model(kvar=5, ar=c(1), ma=c(1,4))

arp <- model$ar.pattern; map = model$ma.pattern
neaus.d <- resid(detr <- lm(detr <- lm(NEaus_ts ~ time(NEaus_ts)), na.action=NULL))
xdata <- matrix(cbind(neaus.d, nino_ts, dmi_ts, tsa_ts, aao_ts), ncol=5) # strip ts attributes
fit <- marima(xdata, ar.pattern=arp, ma.pattern=map, penalty=2)
# resid analysis (not displayed)
innov <- t(resid(fit))

plot.ts(innov[,1])
lines(resid_NE, lwd = 0.75, col = "red", lty = 2)

acf(innov, na.action=na.pass)
pred <- ts(t(fitted(fit))[,1], start=start(NEaus_ts), 
           freq=frequency(NEaus_ts)) + detr$coef[1] + detr$coef[2]*time(NEaus_ts)
plot(pred, lwd = 2, col = "blue")
lines(pred_NE, lwd = 2, col = "red")
#lines(NEaus_ts)
points(NEaus_ts, pch = 16, cex  = 0.5)

residuals_var <- resid(fit)
residuals_var <- residuals_var[,-1]
# Plot ACF and PACF for the first variable's residuals
Acf(residuals_var[,1], main="ACF of Residuals for NE Aus")
Pacf(residuals_var[,1], main="PACF of Residuals for NE Aus")

fit$penalty
fit$ma.estimates
```


```{r SE_aus_VAR}
se_aus_df <- bounded_resp_df[ ,c(2,4,5,6)]

#alter the data to align better, then we can make adjustments to this later
max_resp <- max(nino_df$time)

se_aus_df <- se_aus_df[se_aus_df$time <= max_resp, ]

min_date <- as.Date(min(nino_df$time))

#create a df and ts-object
dates <- se_aus_df$time
SE_aus_tsdf <- data.frame(Date = dates, SE_AUS = se_aus_df$SE_Aus_anomaly_co,
                          Nino = nino_df$nino.anomaly,
                          dmi = dmi_df$dmi.anomaly,
                          tsa = tsa_df$tsa.anomaly,
                          aao = aao_df$aao.anomaly)

SE_aus_zoo <- zoo(SE_aus_tsdf[, -1], order.by = SE_aus_tsdf$Date)

rm(max_resp, min_date)
```

```{r se_var}
var_select <- VARselect(SE_aus_zoo, lag.max = 52, type = "both")

aic_test <- rep(NA, 10)
for (i in 1:10) {
  SE_temp <- vars::VAR(SE_aus_zoo, p = i, type = "both")
  SE_aus_temp <- SE_temp$varresult$SE_AUS
  aic_test[i] <- AIC(SE_aus_temp, k = i + 2)
}

plot(1:10, aic_test, type = "l", xlab = "p", ylab = "AIC", main = "SE Aus VAR(p)")

SE_model <- vars::VAR(SE_aus_zoo, p = 4, type = "both")

SE_aus_model <- SE_model$varresult$SE_AUS

AIC(SE_aus_model, k =2)

summary(SE_model$varresult$SE_AUS)

#change var names and titles below
residuals_var <- residuals(SE_model)

# Plot ACF and PACF for the first variable's residuals
acf(residuals_var[,1], main="ACF of Residuals for SE Aus")
pacf(residuals_var[,1], main="PACF of Residuals for SE Aus")

```

```{r subset_selection}
#NE aus
NE_aus_model <- NE_model$varresult$NE_AUS

summary_neVAR <- summary(NE_aus_model)
coeffs<- summary_neVAR$coefficients 

#threshold:
s <- 1.5
sig_varsNE <- rownames(coeffs)[abs(coeffs[, "t value"]) > s]
sig_varsNE

#SE aus
SE_aus_model <- SE_model$varresult$SE_AUS

summary_seVAR <- summary(SE_aus_model)
coeffs<- summary_seVAR$coefficients 

#threshold:
s <- 1.5
sig_varsSE <- rownames(coeffs)[abs(coeffs[, "t value"]) > s]
sig_varsSE

```


## a. Plot VAR Models

```{r NEAUS_plot}
#NE aus 

NEaus_ts <- ts(bounded_resp_df$NE_Aus_anomaly_co, start = as.Date(min(bounded_resp_df$time)), frequency = 52)
detr <- lm(NEaus_ts ~ time(NEaus_ts))
#t(fitted(NE_aus_model))
test_fit <- fitted(NE_aus_model)
  
pred_NE <- ts(fitted(NE_aus_model), start = start(NEaus_ts), freq = frequency(NEaus_ts))
#+ detr$coef[1] + detr$coef[2]*time(NEaus_ts)

#TODO: fix this plot including ylim and x-val as time.
plot.ts(pred_NE, lwd = 2, col = "blue", ylab = "Anomaly CO", main = "NE Aus Fitted")
#lines(NEaus_ts, lwd = 1, lty = 3)
points(NEaus_ts, pch = 16, cex  = 0.5)

resid_NE <- resid(NE_aus_model)
#resid_NE < NE_aus_model$residuals
plot(resid_NE, pch = 16, ylab = "Residuals", main = "NE Aus Residuals", xlab = "Time")

hist(resid_NE, breaks = 12, col = 'steelblue', 
     main = "NE Aus Residuals",  xlab = "Residuals") 
```

```{r}

SEaus_ts <- ts(bounded_resp_df$SE_Aus_anomaly_co, start = as.Date(min(bounded_resp_df$time)), frequency = 52)
detr <- lm(SEaus_ts ~ time(SEaus_ts))
#t(fitted(NE_aus_model))
test_fit <- fitted(SE_aus_model)
  
pred_se <- ts(fitted(SE_aus_model), start = start(SEaus_ts), freq = frequency(SEaus_ts))
#+ detr$coef[1] + detr$coef[2]*time(SEaus_ts)

#TODO: fix this plot including ylim and x-val as time.
plot(pred_se, lwd = 2, col = "blue", ylab = "Anomaly CO", main = "SE Aus Fitted")
#lines(SEaus_ts)
points(SEaus_ts, pch = 16, cex  = 0.5)

resid_SE <- resid(SE_aus_model)
plot(resid_SE, pch = 16, ylab = "Residuals", main = "SE Aus Residuals", xlab = "Time")

hist(resid_SE, breaks = 12, col = 'steelblue', 
     main = "SE Aus Residuals",  xlab = "Residuals") 
```



# 2. Seasonal Model 

Rename dataframes in below chunks to reflect being the full season.

```{r seasonal_alt}
#set up fire season data

#TODO: create a new col of ne aus wildfire season data,
## if the week is in season_weeks then shift over to new column
## then change 0 in original column

n <- length(ne_aus_df$NE_Aus_anomaly_co)
#add column in 
season_data <- rep(0, length.out = n)
NE_aus_season <- data.frame(ne_aus_df$time, ne_aus_df$NE_Aus_anomaly_co, season_data, 
                            ne_aus_df$week, ne_aus_df$year)
colnames(NE_aus_season) <- c("time", "NE_Aus_anomaly_co", "NE_Aus_season_co",
                             "week", "year")

for (i in 1:n) {
  #get week val first
  temp_week <- NE_aus_season[i, 4]
  if (temp_week %in% season_weeks) {
    NE_aus_season[i, 3] <- NE_aus_season[i, 2]
    NE_aus_season[i, 2] <- 0
  }
}

#TODO: check on how this affects that mean/var of week 52 data.

dates <- NE_aus_season$time
NE_aus_season_df <- data.frame(Date = dates, NE_AUS = NE_aus_season$NE_Aus_anomaly_co,
                          NE_AUS_season = NE_aus_season$NE_Aus_season_co,
                          Nino = nino_df$nino.anomaly,
                          dmi = dmi_df$dmi.anomaly,
                          tsa = tsa_df$tsa.anomaly,
                          aao = aao_df$aao.anomaly)

NE_aus_season_zoo <- zoo(NE_aus_season_df[, -1], order.by = NE_aus_season_df$Date)
```


```{r var_model_seasonal}
var_select <- VARselect(NE_aus_season_zoo, lag.max = 52, type = "both")
aic <- var_select$criteria[1,]
bic <- var_select$criteria[3,]

for (i in 1:10) {
  NE_temp <- vars::VAR(NE_aus_season_zoo, p = i, type = "both")
  NE_aus_temp <- NE_temp$varresult$NE_AUS_season
  print(AIC(NE_aus_temp, k =2))
}

#visual of AIC says p = 7

var_season <-  vars::VAR(NE_aus_season_zoo, p = 7, type = "both")
summary(var_season$varresult$NE_AUS_season)
summary(var_season)


#change var names and titles below
residuals_var <- residuals(var_season)


# Plot ACF and PACF for the first variable's residuals
Acf(residuals_var[,2], main="ACF of Residuals for Variable 2")
Pacf(residuals_var[,2], main="PACF of Residuals for Variable 2")


#plot(1:52, aic, type = "l", xlab = "lag")
#plot(1:52, bic, type = "l", xlab = "lag")

```


```{r season_partition}
#first using var models
early_season <- rep(0, length.out = n)
earlymid_season <- rep(0, length.out = n)
latemid_season <- rep(0, length.out = n)
late_season <- rep(0, length.out = n)
NE_aus_season <- data.frame(ne_aus_df$time, ne_aus_df$NE_Aus_anomaly_co, 
                            early_season, earlymid_season, latemid_season, late_season,
                            ne_aus_df$week, ne_aus_df$year)
colnames(NE_aus_season) <- c("time", "NE_Aus_anomaly_co", 
                             "Early", "Early_Mid", "Late_Mid", "Late",
                             "week", "year")

for (i in 1:n) {
  #get week val first
  temp_week <- NE_aus_season[i, 7]
  if (temp_week %in% early) {
    NE_aus_season[i, 3] <- NE_aus_season[i, 2]
    NE_aus_season[i, 2] <- 0
  }
  if (temp_week %in% early_mid) {
    NE_aus_season[i, 4] <- NE_aus_season[i, 2]
    NE_aus_season[i, 2] <- 0
  }
  if (temp_week %in% late_mid) {
    NE_aus_season[i, 5] <- NE_aus_season[i, 2]
    NE_aus_season[i, 2] <- 0
  }
  if (temp_week %in% late) {
    NE_aus_season[i, 6] <- NE_aus_season[i, 2]
    NE_aus_season[i, 2] <- 0
  }
}

dates <- NE_aus_season$time
NE_aus_season_df <- data.frame(Date = dates, NE_AUS = NE_aus_season$NE_Aus_anomaly_co,
                          NE_Early = NE_aus_season$Early,
                          NE_EarlyMid = NE_aus_season$Early_Mid,
                          NE_LateMid = NE_aus_season$Late_Mid,
                          NE_Late = NE_aus_season$Late,
                          Nino = nino_df$nino.anomaly,
                          dmi = dmi_df$dmi.anomaly,
                          tsa = tsa_df$tsa.anomaly,
                          aao = aao_df$aao.anomaly)

NE_aus_season_zoo <- zoo(NE_aus_season_df[ ,-1], order.by = NE_aus_season_df$Date)

NE_aus_early_zoo <- zoo(NE_aus_season_df[ ,-c(1:2,4:6)], order.by = NE_aus_season_df$Date)
NE_aus_earlyMid_zoo <- zoo(NE_aus_season_df[ ,-c(1:3,5:6)], order.by = NE_aus_season_df$Date)
NE_aus_lateMid_zoo <- zoo(NE_aus_season_df[ ,-c(1:4,6)], order.by = NE_aus_season_df$Date)
NE_aus_late_zoo <- zoo(NE_aus_season_df[ ,-c(1:5)], order.by = NE_aus_season_df$Date)
```


```{r more_refined_VAR}
var_check <- VARselect(NE_aus_season_zoo, lag.max = 52, type = "both")
var_check_early <- VARselect(NE_aus_early_zoo, lag.max = 52, type = "both")
var_check_earlyMid <- VARselect(NE_aus_earlyMid_zoo, lag.max = 52, type = "both")
var_check_lateMid <- VARselect(NE_aus_lateMid_zoo, lag.max = 52, type = "both")
var_check_late <- VARselect(NE_aus_late_zoo, lag.max = 52, type = "both")

var_test <- vars::VAR(NE_aus_season_zoo, p = 12, type = "const")

summary(var_test$varresult$NE_LateMid)

var_latemid <- vars::VAR(NE_aus_lateMid_zoo, p = 2, type = "both")

summary(var_latemid$varresult$NE_LateMid)
```

# 3. Transformed VAR

Transformed using Box-Cox

Use transformed data for a VAR model to see if it looks any better.

```{r trans_data}
#NE_aus
NE_aus_tsdf$NE_AUS <- temp_bounded_resp_df$NE_Aus_anomaly_co

NE_aus_zoo <- zoo(NE_aus_tsdf[, -1], order.by = NE_aus_tsdf$Date)

for (i in 1:10) {
  NE_temp <- vars::VAR(NE_aus_zoo, p = i, type = "both")
  NE_aus_temp <- NE_temp$varresult$NE_AUS
  print(AIC(NE_aus_temp, k =2))
}

NE_model <- vars::VAR(NE_aus_zoo, p = 4, type = "both")

NE_aus_model <- NE_model$varresult$NE_AUS

AIC(NE_aus_model, k =2)

summary(NE_model$varresult$NE_AUS)

#change var names and titles below
residuals_var <- residuals(NE_model)

# Plot ACF and PACF for the first variable's residuals
Acf(residuals_var[,1], main="ACF of Residuals for NE Aus")
Pacf(residuals_var[,1], main="PACF of Residuals for NE Aus")

#SE Aus
SE_aus_tsdf$SE_AUS <- temp_bounded_resp_df$SE_Aus_anomaly_co

SE_aus_zoo <- zoo(SE_aus_tsdf[, -1], order.by = SE_aus_tsdf$Date)


for (i in 1:10) {
  SE_temp <- vars::VAR(SE_aus_zoo, p = i, type = "both")
  SE_aus_temp <- SE_temp$varresult$SE_AUS
  #print(i)
  print(AIC(SE_aus_temp, k =2))
}

SE_model <- vars::VAR(SE_aus_zoo, p = 4, type = "both")

SE_aus_model <- SE_model$varresult$SE_AUS

AIC(SE_aus_model, k =2)

summary(SE_model$varresult$SE_AUS)

#change var names and titles below
residuals_var <- residuals(SE_model)

# Plot ACF and PACF for the first variable's residuals
Acf(residuals_var[,1], main="ACF of Residuals for SE Aus")
Pacf(residuals_var[,1], main="PACF of Residuals for SE Aus")

```

```{r subset_transformed}
#NE Aus
summary_neVAR <- summary(NE_aus_model)
coeffs<- summary_neVAR$coefficients 

#threshold:
s <- 1.5
sig_varsNE <- rownames(coeffs)[abs(coeffs[, "t value"]) > s]
sig_varsNE

#SE aus
summary_seVAR <- summary(SE_aus_model)
coeffs<- summary_seVAR$coefficients 

#threshold:
s <- 1.5
sig_varsSE <- rownames(coeffs)[abs(coeffs[, "t value"]) > s]
sig_varsSE
```


# 4. VARMA models

```{r sEaus}
#residuals have a spike at lag 13 

#test adding ma(13)

```


```{r test varma}
library(astsa) #for cmort data set

model <- define.model(kvar=3, ar=c(1), ma=c(1))

arp <- model$ar.pattern; map = model$ma.pattern
cmort.d <- resid(detr <- lm(cmort~ time(cmort), na.action=NULL))
xdata <- matrix(cbind(cmort.d, tempr, part), ncol=3) # strip ts attributes
fit <- marima(xdata, ar.pattern=arp, ma.pattern=map, means=c(0,1,1), penalty=1)
# resid analysis (not displayed)
innov <- t(resid(fit)); plot.ts(innov); acf(innov, na.action=na.pass)

# fitted values for cmort
pred <- ts(t(fitted(fit))[,1], start=start(cmort), freq=frequency(cmort)) + detr$coef[1] + detr$coef[2]*time(cmort)
plot(pred, ylab="Cardiovascular Mortality", lwd=2, col=4); points(cmort)
```


# 5. Model Validations 





---------------------------------------------
In the below test block we are going to test our VAR model with AIC/BIC but only for predicting the seasonal data.
(and testing of our VARMA model (via library MTS))

Note: the MTS::VARMA also has a refVARMA to get rid of insignificant est to 0

change to marima model

possible steps:
check AIC/BIC for p, q, and threshhold values
for increasing p anf q
for thres 0.5, 1, 1.5 ( adjusted according;y)
output is a pxq matrix where each entry is a tuple of AIC (or BIC) value and the thres value.

```{r test_aicbic}


```


below here is just test blocks, do note use

```{r test_block}
#test ts set up, replace later with real variable names
test_df <- data.frame("time" = nino_df$time, 
                      "ne_aus" = ne_aus_df$NE_Aus_anomaly_co,
                      "nino" = nino_df$nino.anomaly, 
                      "week" = nino_df$week, 
                      "year" = nino_df$year)
test_mat <- matrix(data = c(ne_aus_df$NE_Aus_anomaly_co, nino_df$nino.anomaly),
                   ncol = 2)

colnames(test_mat) <- c("NE_Aus", "Nino")
test_ts <- ts(data = test_mat, start = min_date, frequency = 52)
#test_ts[,1:2]
#print(test_ts) #TODO: get row names to line up with dates (or year + week num)
var_test1 <- VAR(test_ts, p = 5, ic= "AIC", season = 2, type = "trend")
#var_test1
#summary(var_test1)

Date <- seq(as.Date("2020-01-01"), by = "month", length.out = 12)
Date_2 <- test_df$time

xts_test2 <- xts(test_df, order.by =test_df$time)
test_ts2 <- ts(coredata(xts_test2), frequency = 52)
#test_ts2[,1:2]
var_test2 <- VAR(xts_test2, p =5, type = "none")

var_setup <- VARselect(test_ts, lag.max = 52, type = "none")
#summary(var_setup)
#var_setup

#create time series from dfs
#TODO: create a data matrix instead of a df since this isn't really working
nino_df$time <- as.Date(nino_df$time)
xts_test <- xts(nino_df[,-1], order.by = nino_df$time)
#print(xts_test)
ts_test <- as.ts(xts_test, start = min_date)

#ts_test[,1:2]

```

```{r test_block}
#generating an example of data
set.seed(951)

# Generate sample time series data
dates <- seq(as.Date("2015-01-01"), by = "month", length.out = 60)  # 5 years of monthly data
test_data <- seq(as.Date("2011-01-01"), by = "week", length.out = 52)
var1 <- rnorm(60, mean = 100, sd = 10)  # Random data for variable 1
var2 <- rnorm(60, mean = 200, sd = 15)  # Random data for variable 2

# Create the data frame
data_ts <- data.frame(Date = dates, Var1 = var1, Var2 = var2)

# Adding dummy variables for each quarter
data_ts$Q1 <- ifelse(format(dates, "%m") %in% c("01", "02", "03"), 1, 0)
data_ts$Q2 <- ifelse(format(dates, "%m") %in% c("04", "05", "06"), 1, 0)
data_ts$Q3 <- ifelse(format(dates, "%m") %in% c("07", "08", "09"), 1, 0)
data_ts$Q4 <- ifelse(format(dates, "%m") %in% c("10", "11", "12"), 1, 0)

# Convert again to zoo for VAR modeling
data_ts_zoo <- zoo(data_ts[, -1], order.by = data_ts$Date)

# Fit the VAR model including seasonal dummies
var_model_with_dummies <- VAR(data_ts_zoo, p = 2, type = "both")
summary(var_model_with_dummies)
var_model_with_dummies
```

```{r test_case}
test_ts <- data(Canada)
Canada[ ,1]
Canada[ ,1:2]

var_test <- VAR(Canada, p = 2, type = "none")
#VAR(Canada, p = 2, type = "const")
#VAR(Canada, p = 2, type = "trend")
#VAR(Canada, p = 2, type = "both")
```


