---
title: "VAR models"
author: "Ryan Peterson"
date: "2024-05-02"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library}
#for data/date manipulating
suppressMessages(library(lubridate))
suppressMessages(library(tidyverse))
suppressMessages(library(fields)) #test.for.zero

#ts data types
suppressMessages(library(xts))
suppressMessages(library(zoo))

#time series 
suppressMessages(library(vars)) #vector autoregression
suppressMessages(library(forecast)) #spectral methods
```


```{r imported_functions}
#lag_function (requires tidyverse, lubridate)
setwd("~/CO_AUS/Aus_CO-main")
source("lag_function.R")
```


```{r data}
setwd("~/CO_AUS/Aus_CO-main")

#anomaly only response with week data
response_anoms <- read.csv("Data/response_anoms.csv", header = TRUE, stringsAsFactors = FALSE)

#predictors with week data
predictor_anoms <- read.csv("Data/predictor_anoms.csv", header = TRUE, stringsAsFactors = FALSE)

```


```{r data_cleaning}
#TODO: potentially create a function for date bounding
#cleaning data for time series work
week <- 1

response <- response_anoms
predictor <- predictor_anoms

resp_dates <- as_date(response[response$week == week, ]$time)
pred_dates <- as_date(predictor[predictor$week == week, ]$time)

#get minimum prediction date from response data
low_bound <- pred_dates[which(pred_dates == min(resp_dates)) - 1]
#get maximum response date from predictor data
upper_bound <- max(as_date(predictor$time)) + days(7)

#this requires week being set correctly above, that req should be fixed
bounded_pred_df <- predictor[predictor$time >= low_bound & predictor$time <= upper_bound, ]

bounded_resp_df <- response[response$time <= upper_bound, ]#temporarily changed so that each week has the same number of years



rm(response, predictor, low_bound, upper_bound, 
   week, resp_dates, pred_dates)

#outputs to check the boundary dates are correct
min(bounded_pred_df$time)
min(bounded_resp_df$time)

max(bounded_pred_df$time)
max(bounded_resp_df$time)
```


```{r season_setup}
season_weeks <- c(35:52, 1:14)

#early season weeks 35-42
early <- c(35:42)
#early-mid 43-50
early_mid <- c(43:50)
#early-late 52-6
late_mid <- c(51:52, 1:6)
#late 6-14
late <- c(7:14)
```


Outline/Goals:

1. Build simple var model with a response region (NE/SE AUS) and a single climate mode series, with shorter range dependencies (e.g p <= 20).

2.


Notes/Comments
- For the var() function use the season argument to identify the wildfire season along with partitions of the wildfire season.
- Add in seasonality by creating another variable that has the data for that season, basically in season and out of season data.

Questions:


# 1. Var models

Considering NE AUS and Nino

```{r var_data_selection}
#TODO: add in se_aus

#clean data a little bit more
ne_aus_df <- bounded_resp_df[ ,c(2,3,5,6)]
nino_df <- bounded_pred_df[ ,c(2,3,7,8)]
dmi_df <- bounded_pred_df[ ,c(2,4,7,8)]
tsa_df <- bounded_pred_df[ ,c(2,5,7,8)]
aao_df <- bounded_pred_df[ ,c(2,6,7,8)]

#correct for week 53
week_53 <- which(ne_aus_df$week == 53, )
for (k in week_53) {
  ne_aus_df[k-1, 2] <- mean(c(ne_aus_df[k-1, 2], ne_aus_df[k, 2]))
}
rm(k)
ne_aus_df <- ne_aus_df[-week_53, ]


week_53 <- which(nino_df$week == 53, )
test.for.zero(which(tsa_df$week == 53, ), week_53)
test.for.zero(which(dmi_df$week == 53, ), week_53)
test.for.zero(which(aao_df$week == 53, ), week_53)
for (j in week_53) {
  nino_df[j-1, 2] <- mean(c(nino_df[j-1, 2], nino_df[j, 2]))
  dmi_df[j-1, 2] <- mean(c(dmi_df[j-1, 2], dmi_df[j, 2]))
  tsa_df[j-1, 2] <- mean(c(tsa_df[j-1, 2], tsa_df[j, 2]))
  aao_df[j-1, 2] <- mean(c(aao_df[j-1, 2], aao_df[j, 2]))
}
rm(j)
nino_df <- nino_df[-week_53, ]
dmi_df <- dmi_df[-week_53, ]
tsa_df <- tsa_df[-week_53, ]
aao_df <- aao_df[-week_53, ]

#alter the data to align better, then we can make adjustments to this later
min_pred <- min(ne_aus_df$time)
max_resp <- max(nino_df$time)

ne_aus_df <- ne_aus_df[ne_aus_df$time <= max_resp, ]
nino_df <- nino_df[nino_df$time >= min_pred, ]
dmi_df <- dmi_df[dmi_df$time >= min_pred, ]
tsa_df <- tsa_df[tsa_df$time >= min_pred, ]
aao_df <- aao_df[aao_df$time >= min_pred, ]

min_date <- as.Date(min(nino_df$time))

#create a df and ts-object
dates <- ne_aus_df$time
NE_aus_tsdf <- data.frame(Date = dates, NE_AUS = ne_aus_df$NE_Aus_anomaly_co,
                          Nino = nino_df$nino.anomaly,
                          dmi = dmi_df$dmi.anomaly,
                          tsa = tsa_df$tsa.anomaly,
                          aao = aao_df$aao.anomaly)

NE_aus_zoo <- zoo(NE_aus_tsdf[, -1], order.by = NE_aus_tsdf$Date)

rm(min_pred, max_resp, min_date)
```


```{r just_test_it_out}
#no seasonality
var_select <- VARselect(NE_aus_zoo, lag.max = 52, type = "both")
aic <- var_select$criteria[1,]
bic <- var_select$criteria[3,]

NE_model <- VAR(NE_aus_zoo, p = 3, type = "both")

summary(NE_model$varresult$NE_AUS)

plot(1:52, aic, type = "l", xlab = "lag")
plot(1:52, bic, type = "l", xlab = "lag")
```


```{r seasonal_alt}
#set up fire season data

#TODO: create a new col of ne aus wildfire season data,
## if the week is in season_weeks then shift over to new column
## then change 0 in original column

n <- length(ne_aus_df$NE_Aus_anomaly_co)
#add column in 
season_data <- rep(0, length.out = n)
NE_aus_season <- data.frame(ne_aus_df$time, ne_aus_df$NE_Aus_anomaly_co, season_data, 
                            ne_aus_df$week, ne_aus_df$year)
colnames(NE_aus_season) <- c("time", "NE_Aus_anomaly_co", "NE_Aus_season_co",
                             "week", "year")

for (i in 1:n) {
  #get week val first
  temp_week <- NE_aus_season[i, 4]
  if (temp_week %in% season_weeks) {
    NE_aus_season[i, 3] <- NE_aus_season[i, 2]
    NE_aus_season[i, 2] <- 0
  }
}

dates <- NE_aus_season$time
NE_aus_season_df <- data.frame(Date = dates, NE_AUS = NE_aus_season$NE_Aus_anomaly_co,
                          NE_AUS_season = NE_aus_season$NE_Aus_season_co,
                          Nino = nino_df$nino.anomaly,
                          dmi = dmi_df$dmi.anomaly,
                          tsa = tsa_df$tsa.anomaly,
                          aao = aao_df$aao.anomaly)

NE_aus_season_zoo <- zoo(NE_aus_season_df[, -1], order.by = NE_aus_season_df$Date)
```


```{r var_model_seasonal}
var_select <- VARselect(NE_aus_season_zoo, lag.max = 52, type = "both")
aic <- var_select$criteria[1,]
bic <- var_select$criteria[3,]

plot(1:52, aic, type = "l", xlab = "lag")
plot(1:52, bic, type = "l", xlab = "lag")

```


below here is just test blocks, do note use

```{r test_block}
#test ts set up, replace later with real variable names
test_df <- data.frame("time" = nino_df$time, 
                      "ne_aus" = ne_aus_df$NE_Aus_anomaly_co,
                      "nino" = nino_df$nino.anomaly, 
                      "week" = nino_df$week, 
                      "year" = nino_df$year)
test_mat <- matrix(data = c(ne_aus_df$NE_Aus_anomaly_co, nino_df$nino.anomaly),
                   ncol = 2)

colnames(test_mat) <- c("NE_Aus", "Nino")
test_ts <- ts(data = test_mat, start = min_date, frequency = 52)
#test_ts[,1:2]
#print(test_ts) #TODO: get row names to line up with dates (or year + week num)
var_test1 <- VAR(test_ts, p = 5, ic= "AIC", season = 2, type = "trend")
#var_test1
#summary(var_test1)

Date <- seq(as.Date("2020-01-01"), by = "month", length.out = 12)
Date_2 <- test_df$time

xts_test2 <- xts(test_df, order.by =test_df$time)
test_ts2 <- ts(coredata(xts_test2), frequency = 52)
#test_ts2[,1:2]
var_test2 <- VAR(xts_test2, p =5, type = "none")

var_setup <- VARselect(test_ts, lag.max = 52, type = "none")
#summary(var_setup)
#var_setup

#create time series from dfs
#TODO: create a data matrix instead of a df since this isn't really working
nino_df$time <- as.Date(nino_df$time)
xts_test <- xts(nino_df[,-1], order.by = nino_df$time)
#print(xts_test)
ts_test <- as.ts(xts_test, start = min_date)

#ts_test[,1:2]

```

```{r test_block}
#generating an example of data
set.seed(951)

# Generate sample time series data
dates <- seq(as.Date("2015-01-01"), by = "month", length.out = 60)  # 5 years of monthly data
test_data <- seq(as.Date("2011-01-01"), by = "week", length.out = 52)
var1 <- rnorm(60, mean = 100, sd = 10)  # Random data for variable 1
var2 <- rnorm(60, mean = 200, sd = 15)  # Random data for variable 2

# Create the data frame
data_ts <- data.frame(Date = dates, Var1 = var1, Var2 = var2)

# Adding dummy variables for each quarter
data_ts$Q1 <- ifelse(format(dates, "%m") %in% c("01", "02", "03"), 1, 0)
data_ts$Q2 <- ifelse(format(dates, "%m") %in% c("04", "05", "06"), 1, 0)
data_ts$Q3 <- ifelse(format(dates, "%m") %in% c("07", "08", "09"), 1, 0)
data_ts$Q4 <- ifelse(format(dates, "%m") %in% c("10", "11", "12"), 1, 0)

# Convert again to zoo for VAR modeling
data_ts_zoo <- zoo(data_ts[, -1], order.by = data_ts$Date)

# Fit the VAR model including seasonal dummies
var_model_with_dummies <- VAR(data_ts_zoo, p = 2, type = "both")
summary(var_model_with_dummies)
var_model_with_dummies
```

```{r test_case}
test_ts <- data(Canada)
Canada[ ,1]
Canada[ ,1:2]

var_test <- VAR(Canada, p = 2, type = "none")
#VAR(Canada, p = 2, type = "const")
#VAR(Canada, p = 2, type = "trend")
#VAR(Canada, p = 2, type = "both")
```


